\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-03 canonicitymodel.tex
Small fixes throughout metatheory section

Now we sketch out the big picture of the proof. The canonicity model $"Tm"^C~\Gamma^C~"T"^C$ will map each syntax piece $("t : Tm Γ T") \mapsto "(t, tₚ) :" "Tm"^C~\Gamma^C~"T"^C = \sum~("t : Tm Γ T")."Tm"_p~\Gamma^C~"T"^C~"t"$ to a dependent pair\footnote{This $"Tm"_p$ is usually called dependent model.}, of which the first part will be the same as the input syntax. Using this mapping on closed boolean term $"t : Tm ⋅ "\mathbb{B}$ will give us $"Tm"_p~\cdot^C~\mathbb{B}^C~"t"$, which will unfolds to our final goal $"t" \equiv "tt" + "t" \equiv "ff"$. Roughly, this can read as ``every closed boolean term is reducible''.  Thus once we can ``propagate'' $(\cdot)^C$ into every syntactic piece, and construct this model, we are done. In our (fake) Agda formulation, we use $(\cdot)_2$ to replace the notation of $(\cdot)^C$ because we consider most parts of this model are actually a dependent pair.

Another helpful analogy of the canonicity model is that $"Tm"_p~\Gamma^C~"T"^C~"t"$ can be understood as $"t" \in \{$ Reducible Terms of type $T$ (and $T_p$) in context $\Gamma$ (and $\Gamma_p$) $\}$. The latter is more conventional and the former can be understood as type theoretic encoding of the latter\footnote{We usually use predicate to encode "belong to" relation in type theory}. From this perspective, this part of the canonicity model is quite similar to how conventional (proof-irrelevant) logical relation is carried out~\cite{skorstengaard2019introduction}---using the terminology of the conventional formulation, $\Gamma_p$ will be the set of substitution into closed reducible terms, and we will prove $t[\gamma] \in T_p$, a set of closed reducible term of type $T$ when $\gamma \in \Gamma_p$.

>>>>

Now we sketch out the big picture of the proof. The canonicity model $"Tm"^C~\Gamma^C~"T"^C$ will map each syntax piece $("t : Tm Γ T") \mapsto "(t, tₚ) :" "Tm"^C~\Gamma^C~"T"^C = \sum~("t : Tm Γ T")."Tm"_p~\Gamma^C~"T"^C~"t"$ to a dependent pair\footnote{This $"Tm"_p$ is usually called dependent model.}, of which the first part will be the same as the input syntax. Using this mapping on closed boolean term $"t : Tm ⋅ "\mathbb{B}$ will give us $"Tm"_p~\cdot^C~\mathbb{B}^C~"t"$, which will unfolds to our final goal $"t" \equiv "tt" + "t" \equiv "ff"$. Roughly, this can read as ``every closed boolean term is reducible''.  Thus once we can ``propagate'' $(\cdot)^C$ into every syntactic piece, and construct this model, we are done. In our (pseudo) Agda formulation, we use $(\cdot)_2$ to replace the notation of $(\cdot)^C$ because we consider most parts of this model are actually a dependent pair.

Another helpful analogy of the canonicity model is that $"Tm"_p~\Gamma^C~"T"^C~"t"$ can be understood as $"t" \in \{$ Reducible Terms of type $T$ (and $T_p$) in context $\Gamma$ (and $\Gamma_p$) $\}$. The latter is more conventional and the former can be understood as type theoretic encoding of the latter\footnote{We usually use predicate to encode ``belong to''/``is element of'' relation in type theory}. From this perspective, this part of the canonicity model is quite similar to how conventional (proof-irrelevant) logical relation is carried out~\cite{skorstengaard2019introduction}---using the terminology of the conventional formulation, $\Gamma_p$ will witness the set of substitution into closed reducible terms of types $\Gamma$. Then we will prove $t[\gamma] \in T_p$---a set of closed reducible term of type $T$---when $\gamma \in \Gamma_p$.

### 2022-07-03 coqimpl.tex
Small fixes throughout metatheory section

This raise an interesting question: what propositions are conservative extension? For example, apparently we know injectiveness of constructor, (i.e. \mintinline{Coq}{STLC.tt ≠ STLC.ff}) must be conservative because it holds for all extension of the inductive type "STLC.tm". But is there a ``best'' proposition (i.e. the proposition that everything the proposition derives are exactly conservative extension)?

@@

Even though we cannot prove propositional partial recursor is the best,
the above "reflect" function implies that propositional partial
recursor is a good \textit{extensional characterization} of (non-indexed) extensible inductive type (with certain constraints): because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

>>>>

This raise an interesting question: what propositions are conservative extension? For example, apparently we know injectiveness of constructor, (i.e. \mintinline{Coq}{STLC.tt ≠ STLC.ff}) must be conservative because it holds for all extension of the inductive type "STLC.tm". But is there a \textbf{``best'' proposition} (i.e. the proposition that everything the proposition derives are exactly conservative extension)?

@@

Even though we cannot prove propositional partial recursor is \textit{the best},
the above "reflect" function implies that propositional partial
recursor is a good \textit{extensional characterization} of (non-indexed) extensible inductive type (with certain constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

### 2022-07-03 metatheory.tex
Small fixes throughout metatheory section

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic a type theoretical framework (like Agda), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Our work can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We base our formulation of the syntax in this type theory instead of a more conventional and informal representation, becase (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a proof because proof checking is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real theorem prover so that the reader can have an idea about how mechanized formalization can be carried out.

>>>>

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic a type theoretical framework (like Agda), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Our work can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We base our formulation of the syntax in this type theory instead of a more conventional and informal representation, because (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a proof because proof checking then is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real theorem prover so that the reader can have an idea about how mechanized formalization can be carried out.

### 2022-07-03 standardmodel.tex
Small fixes throughout metatheory section

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from~\cite{altkap2016,kaposi2017type, kaposi2019gluing} and their formalization, where they formally resort to the concept of \textit{algebra/model} of QIIT, here we only consider inductively interpreting each syntax piece into naive set theory that is also respecting judgemental equalities\footnote{In our appendix, we still interpret into Agda component to be more precise and less error-prone, also we need the inductive facility from Agda.} The main idea of the proof is that: 1. we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; 2. we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$\footnote{Actually our 1 is acting exactly as an interpretation function, and in our agda formulation we only do 1 because 2 is automatic; here we make 2 explicit to make our formulation more accessible and uniform}, and also we omit the verification of equality (i.e. the judgementally-equal syntax needs to be mapped to the same data for well-definedness).

Note that, since every piece is intrinsic well-typed, $\denotesS{t}$ means the same as $\denotesS{\goodTerm{\Gamma}{t}{T}}$ (i.e. every term is tracking its context and type)

@@

Notice that $\denotesS{\goodSig{\Gamma}{s}{n}}$ and $\denotesS{\goodType{\Gamma}{\cC\sigma}{i}}$ is mutual recursively defined together---because $\goodSealS{\Gamma}{f}{\sigma}{A}$ relies on $\cC$.

@@

Since all the compatibility lemmas hold (by simply changing $\vdash$ into $\models$ in the syntactic rules then we would have all the compatibility lemmas\footnote{Apparently, in our (fake) Agda formulation, we don't need this extra step.}), we have constructed a model.
Finally, notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot$ which exactly means $\denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.

The full model formulated in (fake) Agda syntax can be found in the appendix.

>>>>

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from~\cite{altkap2016,kaposi2017type, kaposi2019gluing} and their formalization, where they formally resort to the concept of \textit{algebra/model} of QIIT, here we only consider inductively interpreting each syntax piece into naive set theory that is also respecting judgemental equalities\footnote{In our appendix, we still interpret into Agda component to be more precise and less error-prone, also we need the inductive facility from Agda.}

The main idea of the proof is that: (1) we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; (2) we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$\footnote{Actually our 1 is acting exactly as an interpretation function, and in our (pseudo) Agda formulation we only do 1 because 2 is automatic; here we make 2 explicit to make our formulation more accessible and uniform}, and also we omit the verification of equality (i.e. the judgementally-equal syntax needs to be mapped to the same data for well-definedness).

The main intuition of the constructed consistency model is that: (1) since we have used sigma type to translate the linkage, using the model for sigma type for linkage will work; (2) directly using the inductive facility of (pseudo)-Agda to justify the intro and elim rule of our W-type.

Note that, since every piece is intrinsic well-typed, $\denotesS{t}$ means the same as $\denotesS{\goodTerm{\Gamma}{t}{T}}$ (i.e. every term is tracking its context and type).\footnote{And our (pseudo) Agda formulation is well-typed as well. Latex-formulation should be well-typed but we discarded all the type information in the paper.}

@@

Notice that $\denotesS{\goodSig{\Gamma}{s}{n}}$ and $\denotesS{\goodType{\Gamma}{\cC\sigma}{i}}$ is mutual recursively defined together---because \\ $\goodSealS{\Gamma}{f}{\sigma}{A}$ relies on $\cC$.

@@

Since all the compatibility lemmas hold (by simply changing $\vdash$ into $\models$ in the syntactic rules then we would have all the compatibility lemmas\footnote{Apparently, in our (pseudo) Agda formulation, we don't need this extra step.}), we have constructed a model.
Finally, notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot$ which exactly means $\denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.

The full model formulated in (pseudo) Agda syntax can be found in the appendix.

### 2022-07-03 syn-translate.tex
Small fixes throughout metatheory section

The implementation described in \cref{sec:coqimpl} can be formalized in our metatheory. The described compilation strategy can be considered as a \textit{syntactical translation}, a function mapping from our syntactical (QIIT-)model to a subpart of our syntactical (QIIT-)model \textit{without linkages}.\footnote{Here, syntactical QIIT-model can be roughly understood as just AST, with equational rules acting like quotient. So this mapping is mapping AST to AST, respecting quotient/judgemental equality.} This subpart is the standard MLTT with $\goodWSig{}{}{}$, a (non-conventional) W-type and a recursor that using sigma type to aggregate the handlers.

@@

Ultimately, we have constructed a model for our QIIT-syntax using only subparts of our QIIT-syntax -- without these linkages. This also justify the intuition -- a linkage is actually a module \textit{with universal quantification wrapping its fields to achieve late binding}.

It is also possible to compile away $\goodWSig{\Gamma}{w}{n}$ but the recursor will always require a sigma type as handler and thus the elimination rule for our W-type is still non-conventional.

After syntactical translation, we can directly have consistency and canonicity \textit{if we assume the consistency and canonicity of the subpart}. However, due to our non-conventional formulation of the inductive type, it is better we don't impose such assumption and directly give a consistency and canonicity proof.

>>>>

The implementation described in \cref{sec:coqimpl} can be formalized in our metatheory. The described compilation strategy can be considered as a \textit{syntactical translation}, a function mapping from our syntactical (QIIT-)model to a ``subpart''\footnote{Not exactly subpart---the only difference lies in the "Wrec", in our metatheory, it is constructed using linkage; in ``subpart'', it will be using sigma types (translation of the linkage).} of our syntactical (QIIT-)model \textit{without linkages}.\footnote{Here, syntactical QIIT-model can be roughly understood as just AST, with equational rules acting like quotient. So this mapping is mapping AST to AST, respecting quotient/judgemental equality.} This ``subpart'' is the standard MLTT with inductive type signature, a (non-conventional) W-type and a recursor that using sigma type to aggregate the handlers.

@@

Ultimately, we have constructed a model for our QIIT-syntax using only a ``subpart'' of our QIIT-syntax -- without these linkages. It is also possible to compile our $\goodWSig{\Gamma}{w}{n}$ and W-type into a more conventional formulation, but the recursor will always require a sigma type as handler and thus the elimination rule for our W-type is still non-conventional.

This translation justifies the intuition -- a linkage is actually a module \textbf{with universal quantification wrapping its fields to achieve late binding}.

After syntactical translation, we can directly have consistency and canonicity \textit{if we assume the consistency and canonicity of the ``subpart''}. However, due to our non-conventional formulation of the inductive type, it is better we don't impose such assumption and directly give a consistency and canonicity proof.
